---
title: "Tracing Trees"
subtitle: "Visualizing the Architecture of Random Forests with Extensions of Trace Plots"
author: 
  - "<br>"
  - ".light_green[Katherine Goode]"
  - ".light_green[Presented at ISU Graphics Group]"
date: ".light_green[April 1, 2021] <br> <br> <br> <br> .medium[.white[Code for creating slides available [here](https://github.com/goodekat/presentations/tree/master/2021-isugg-trace-plots)]]"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{css, echo = FALSE}
.tiny{font-size: 30%}
.small{font-size: 50%}
.medium{font-size: 75%}
.left-code {
  color: #777;
  width: 37%;
  height: 92%;
  float: left;
}
.right-plot {
  width: 60%;
  float: right;
  padding-left: 1%;
}
```

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)
style_mono_accent(
  white_color = "#f7f7e8",
  base_color = "#557174",
  header_font_google = google_font("Josefin Sans"),
  title_slide_background_color = "#557174",
  inverse_background_color = "#557174",
  text_font_google = google_font("Montserrat", "300", "300i"),
  code_font_google = google_font("Fira Mono"),
  text_slide_number_color = "black",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.25rem",
  colors = c(
    cream = "#f7f7e8",
    light_green = "#c7cfb7",
    medium_green = "#9dad7f",
    dark_green = "#557174"
  ),
  link_color = "#9dad7f",
  code_highlight_color = "#9dad7f"
)
```

# Overview

- Intro to random forests 

- Background on trace plots

- Extending Trace Plots

  - Visualizing the canopy
  
  - Summary trees with variability
  
  - Clusters of trees
  
  - Linking to other plots

- Limitations and going forward

---

class: inverse, center, middle

# Intro to random forests

---

class: inverse, center, middle

# Background on trace plots

---

class: inverse, center, middle

# Extending trace plots

---

# Visualizing the canopy

---

# Summary trees with variability

---

# Clusters of trees

---

# Linking to other plots

---

class: inverse, center, middle

# Limitations and going foward
